### 2010-02-01 trunk/deliverables/D4_3_2/fast_d4_3_2.tex
adds more content and integrates Albert's contribution
git-svn-id: https://svn.forge.morfeo-project.org/fast-fp7project@494 d8b0c59b-6544-0410-94d9-83a5658a058e

\newcommand\deliverableNumber{D4.3.2}
\newcommand\deliverableTitle{Mechanisms for Gadget-Service Connections and Gadget Functionality}
\newcommand\deliverableTitleShort{Mechanisms for Gadget-Service Connections}
\newcommand\workpackageNumber{4}
\newcommand\workpackageTitle{Visual composition of screen-flow resources and interoperability with back-end Web Services}
\newcommand\authorOne{Ismael Rivera, NUIG}
\newcommand\authorTwo{Author 2}
\newcommand\authorThree{Author 3}
\newcommand\authorFour{Author 4}
\newtheorem{example}{\emph{Example}}

@@

%\begin{minipage}[p]{15cm}
	\begin{center}
		\includegraphics{images/FAST_logo}\\
		\vspace{1cm}
		{\LARGE{\sffamily \emph{FAST AND ADVANCED STORYBOARD TOOLS}}}\\
		\vspace{0.5cm}
		{\LARGE \sffamily \emph{FP7-ICT-2007-1-216048}}\\
		\vspace{0.5cm}
		{\LARGE \sffamily \emph{http://fast.morfeo-project.eu}}\\
		\vspace{4cm}
		{\LARGE \sffamily \textbf{Deliverable \deliverableNumber}}\\
		\vspace{0.5cm}
		{\LARGE \sffamily \textbf{\deliverableTitle}}\\
		\vspace{2cm}
		{\large \sffamily \authorOne}\\
%		{\large \sffamily \authorTwo}\\
		\vspace{0.5cm}
		\vfill
		{\large \sffamily Date: 26/02/2010}\\
		\vspace{1cm}
		{\sffamily FAST is partially funded by the E.C. (grant code: FP7-ICT-2007-1-216048).}

@@

\begin{small}
\begin{tabular}{|l|l|l|p{7.5cm}|}
\hline
\rowcolor{fast@lightgrey}\textcolor{white}{\textbf{Rev. No.}} &
                            \textcolor{white}{\textbf{Date}} &
                            \textcolor{white}{\textbf{Author (Partner)}} &
							\textcolor{white}{\textbf{Change description}}\\ \hline
%0.1 & 15.12.2008 & Knud M\"oller (NUIG) & template modelled \\ \hline
%0.2 & 15.02.2009 & Knud M\"oller (NUIG) & content moved from wiki after internal review \\ \hline
1.0 & 26.02.2010 & Ismael Rivera (NUIG) & Final version ready for external review \\ \hline
\end{tabular}
\end{small}

@@

These back-end services cannot be directly used; hence they need to be encapsulated in what in the FAST platform is called Resource Adapters. The application of semantics to the back-end, through the corresponding Resource Adapters, and front-end building blocks assures a powerful instrument in the task of building new gadgets, improving the search, and enhancing the connection among the different building blocks which compose a gadget.

@@

\begin{tabular}
	{| >{\columncolor{fast@lightgrey}}p{3.25cm}|p{1.4cm}|p{3.28cm}|p{1.6cm}|p{3.29cm}|}
	\hline
	\textcolor{white}{\textbf{Delivery data}} & {\textbf{Due date}} & {28/02/2009} & {\textbf{Submitted}} & {27/02/2009}\\ \hline
	\textcolor{white}{\textbf{Status}} & \multicolumn{2}{l|}{} & \multicolumn{2}{l|}{final}\\ \hline
	\textcolor{white}{\textbf{Dissemination Level}} & \multicolumn{4}{l|}{Public $\boxtimes$ / Consortium $\square$}\\ \hline
	\textcolor{white}{\textbf{Short description of contents}} & \multicolumn{4}{p{10.85cm}|}{D4.3.2 is...}\\ \hline
	\textcolor{white}{\textbf{Authors}} & \multicolumn{4}{l|}{\authorOne}\\
%	{} & \multicolumn{4}{l|}{}\\ 
%	{} & \multicolumn{4}{l|}{}\\ 
%	{} & \multicolumn{4}{l|}{}\\
	\hline
	\textcolor{white}{\textbf{Deliverable Owner}} & \multicolumn{2}{l|}{\authorOne} & \textbf{email} & {ismael.rivera@deri.org} \\ \cline{4-5}
	\textcolor{white}{\textbf{(Partner)}} & \multicolumn{2}{l|}{} & \textbf{phone} & {+353 91 ******} \\ \hline
	\textcolor{white}{\textbf{Keywords}} & \multicolumn{4}{p{10.85cm}|}{FAST, web services, WSDL, REST, SOAP, WSMO}\\ \hline
\end{tabular}
\end{small}
\newpage

@@

This deliverable is structured as follows. Section ~\ref{sec:introduction} states the goal, scope and structure of the document. Section ~\ref{sec:web_services_wrapping} describes what a service wrapper is, how the platform is able to interact with web services, how they these wrappers are build and then discovered to be reused in any gadget. Finally, Section ~\ref{sec:restful_web_services_wrapper_tool} defines the service wrapper tool built to facilitate FAST users to create Resource Adapters of RESTful web services.

@@

\clearpage
\section{Web Services Wrapping} % (fold)
\label{sec:web_services_wrapping}

@@

Therefore, a complex gadget in terms of FAST is aimed to provide a functional access from a graphical user interface to a set of services (SOAP or REST-based Web services) and data sources (Atom/RSS feeds). These services and data sources need to be modelled and encapsulated within the platform in order to allow the discovery and use of them by other components (i.e. forms and operators). This is called Resource Adapter in the FAST platform. Figure ~\ref{fig:complex_gadget_architecture} illustrates their position and interaction with the rest of the building blocks within a gadget.

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{images/ComplexGadgetArchitecture.pdf}
    \caption{Complex Gadget Architecture}
    \label{fig:complex_gadget_architecture}
  \end{center}
\end{figure}

Once a brief background has been shown, the following sections explain how a user, in this case a resource developer, can build Resource Adapters, and how another user, a screen developer for instance, will discover and connect them while creating a screen. Hence, these three phases are called: building, discovery and connection.

@@

@@

REST is a term to describe an architecture style, not a standard, of networked systems. The acronym REST stands for Representational State Transfer. REST-based or RESTful web services \cite{Fielding2000} are created identifying all of the conceptual entities or resources which want to be exposed as services. Those entities or resources should be nouns, not verbs (orders, tickets, etc.). Then, the interaction with those resources is made by convention using HTTP verbs such as GET, POST, PUT or DELETE, in order to retrieve, create, modify or delete them.

REST is lightweight (not a lot of extra xml markup, human readable results), but unlike SOAP-based web services, which have a standard vocabulary, and commonly used, to describe the web service interface through WSDL, RESTful web services are currently not formally described most of the times. For a service consumer to understand the context and content of the data that must be sent to and received from the service, both the service consumer and service producer must have an out-of-band agreement. This takes the form of documentation, sample code, and an API that the service provider publishes for developers to use. For example, the many web-based services available from Google, Yahoo, Flickr, Amazon, and so on have accompanying artefacts describing how to consume the services. This style of documenting REST-based web services is fine for use by developers, but it averts tools from programmatically consuming such services and generating artefacts specific to programming languages, as a web service described using WSDL allows. Nevertheless, Web Application Description Language (WADL) attempts to resolve some of these issues by providing a means to describe services in terms of schemas, HTTP methods, and the request or response structures exchanged, but this language is not widely adopted yet by RESTful web service developers.

To permit a high number of REST-based web services to be integrated to the FAST platform, the approach taken is from a manual development perspective. There is no need of a formal document such as WADL defining the web service, for this reason, building service wrappers for these services involve the correct understanding of the service by a human being and a tool to facilitate this task. This tool is explained in detail in Section ~\ref{sec:restful_web_services_wrapper_tool}.

@@

SOAP-based web Services or "Heavyweight Web Services" use Extensible Markup Language (XML) \cite{XML1.1} messages that follow the Simple Object Access Protocol (SOAP) standard \cite{SOAP1.2} and have been popular with traditional enterprise, usually relying on HTTP for message negotiation and transmission. In such services, there is often a machine-readable description of the operations offered by the service written in a Web Services Description Language (WSDL) document. Hence, the advantage of using WSDL is that it can be programmatically processed.

@@

From a specific WSDL definition, a set of methods or operations can be easily extracted which encloses a set of inputs and outputs. These operations would be transformed into actions, and the inputs and outputs would be used to define Resource Adapter pre/postconditions. With the use of this information along with the XML Schema defining the types, a Resource Adapter can be semantically defined.

Moreover, there are several frameworks to facilitate the programmatic use of these formal definitions of web services. These frameworks shall be use to transform a WSDL definition into an executable Resource Adapter ready to use inside a gadget.

@@

A web service in WSMO defines one and only one capability. The capability of a web service defines its functionality in terms of pre/postconditions, assumptions and effects. A web service capability is defined by specifying the following elements: non-functional properties, imported ontologies, used mediators, shared variables, precondition, postcondition, assumption, and effect.

Basically, the web service will offers to a client a postcondition when some conditions are met in the information space (precondition). Effects and assumptions can stay out of this study, since they will not be taken into account by any of the building blocks of a gadget.

NEED TO BE FINISHED

@@

\subsection{Discovering} % (fold)
\label{sub:discovering}

In the building phase, every Resource Adapter is created sharing a common structure for screen components, such as forms, operators and resources. Hence, it will have a set of actions which will contain a set of preconditions to satisfy in order to be executed, and after its execution, it may produce any of the conditions of their postconditions.

@@

In these bases, the mechanism is relatively similar to the screen recommender, used to find the best screens to create a screen flow. Once more, it is based on the pre/postconditions of a set of building blocks.

@@

@@

In FAST, the wrapper tool will be in charge of the Resource Adapters' building phase. At the moment, just supports RESTful web services. The building phase for this type of web services is done in two steps: a first step will be in charge of the construction of a service request and a second step will analyse the response got from the execution of the service, allowing the extraction of facts contained in that response and mapping them to domain-specific concepts from the ontologies used within FAST by any building block.

@@

To achieve this, service wrappers will be handled as pseudo-screens with preconditions and postconditions (inputs and outputs). Therefore, these precondition ports might be used to determine values to parameters like query keywords coming from a building block inside the screen the service is placed or even an external screen as a screen precondition.

To define the preconditions and postconditions of a new service wrapper the FAST service wrapper tool provides a form that allows the editing of these entries, cf. Figure \ref{fig:construct_pre_post_conditions}\footnote{This screen dump is a combination of two different FAST tool screen dumps that show how the final tool will look like. We add a real screen dump ASAP.}. Note, we allow to edit example values for the preconditions. These example values may be used to test the service wrapper.

\begin{figure}
  \begin{center} \includegraphics[width=\linewidth]{images/ServiceWrapperToolGVSWithPortDefinitions.jpg}
    \caption{Configuring pre- and postconditions of a service wrapper (photoshoped screen dump)}
    \label{fig:construct_pre_post_conditions}
  \end{center}
\end{figure}

@@

Figure ~\ref{fig:construct_service_request} shows the screen to construct the service requests.

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{images/ServiceWrapperToolGVSWithRequestExample.jpg}
    \caption{Constructing the service request URL and its parameters (photoshoped screen dump)}
    \label{fig:construct_service_request}
  \end{center}
\end{figure}

In the top input field of Figure ~\ref{fig:construct_service_request}, the user may drop an example http request taken e.g. form the service documentation e.g. from \cite{eBayShoppingAPIs}. The tool analyses the example request and in the middle of the screen a form for editing the request parameters is provided. In our example, the user has connected the \textit{QueryKeywords} parameter with the precondition \textit{search\_key} by adding a corresponding reference to the value field of that parameter. In addition, we have retrieved an access key which has been entered as value for the \textit{appid} parameter.

Below the request paramter editing form of Figure ~\ref{fig:construct_service_request}, a \textit{Send Request} button allows to validate the constructed service by sending it to the specified service address (via a server relay). Then the placeholders for preconditions are replaced with example values and the resulting http request is shown below the parameter form. In addition, the request is send and the response is shown on the bottom of that page. This gives the user a fast feedback whether the constructed request works as desired.

\subsubsection{Limitations} % (fold)
\label{ssub:limitations}

It is worth pointing out that currently the wrapper tool is able to construct input ports for the wrappers using just basic types. However, every building block accepts any (complex) concept as a pre/postcondition, so the wrapper tool needs to be adapted to permit any concept as input or precondition, extracting the required text to construct the request. This is planned for the third year.

% subsubsection limitations (end)

@@

Once the service request is constructed and sent to the service provider, it will send back a response. This response message could be serialize in any format, though the most common formats used nowadays are XML or JSON among others. To continue the example started in the previous section, the response of the Ebay Shopping service will be in XML as specified in the request, which is the format supported by the wrapper tool.

\subsubsection{Translation XML into Facts} % (fold)
\label{ssub:translation_xml_into_facts}

Figure ~\ref{fig:response_service_execution} shows the data tranformation tab of the wrapper tool.

>>>>

% This is useful for switching in-line comments on and off!
\usepackage{ifthen}
\newboolean{showcomments}
\setboolean{showcomments}{true}
\ifthenelse{\boolean{showcomments}}
{
	\definecolor{Red}{rgb}{1,0,0}
	\newcommand{\note}[1]{\begin{scriptsize}{{\textsf{\textbf{{\textcolor{Red}{[[#1]]}}}}}}\end{scriptsize}}
	\definecolor{Orange}{rgb}{1,0.5,0}
	\newcommand{\todo}[1]{\textsf{\textbf{\textcolor{Orange}{[[TODO: #1]]}}}}
    \definecolor{Green}{rgb}{0,1,0}
    \newcommand{\commentreply}[1]{\begin{scriptsize}{\textsf{\textbf{\textcolor{Green}{[[#1]]}}}}\end{scriptsize}}
} {
	\newcommand{\note}[1]{\textbf{}}
	\newcommand{\todo}[1]{}
    \newcommand{\commentreply}[1]{}
}

\newcommand\deliverableNumber{D4.3.2}
\newcommand\deliverableTitle{Mechanisms for Gadget-Service Connections and Gadget Functionality}
\newcommand\deliverableTitleShort{Mechanisms for Gadget-Service Connections}
\newcommand\workpackageNumber{4}
\newcommand\workpackageTitle{Visual composition of screen-flow resources and interoperability with back-end Web Services}
\newcommand\authorOne{Ismael Rivera, NUIG}
\newcommand\authorTwo{Albert Z\"undorf, UniKassel}
\newcommand\authorThree{Author 3}
\newcommand\authorFour{Author 4}
\newtheorem{example}{\emph{Example}}

@@

%\begin{minipage}[p]{15cm}
	\begin{center}
		\includegraphics{images/FAST_logo}\\
		\vspace{1cm}
		{\LARGE{\sffamily \emph{FAST AND ADVANCED STORYBOARD TOOLS}}}\\
		\vspace{0.5cm}
		{\LARGE \sffamily \emph{FP7-ICT-2007-1-216048}}\\
		\vspace{0.5cm}
		{\LARGE \sffamily \emph{http://fast.morfeo-project.eu}}\\
		\vspace{4cm}
		{\LARGE \sffamily \textbf{Deliverable \deliverableNumber}}\\
		\vspace{0.5cm}
		{\LARGE \sffamily \textbf{\deliverableTitle}}\\
		\vspace{2cm}
		{\large \sffamily \authorOne}\\
		{\large \sffamily \authorTwo}\\
		\vspace{0.5cm}
		\vfill
		{\large \sffamily Date: 26/02/2010}\\
		\vspace{1cm}
		{\sffamily FAST is partially funded by the E.C. (grant code: FP7-ICT-2007-1-216048).}

@@

\begin{small}
\begin{tabular}{|l|l|l|p{7.5cm}|}
\hline
\rowcolor{fast@lightgrey}\textcolor{white}{\textbf{Rev. No.}} &
                            \textcolor{white}{\textbf{Date}} &
                            \textcolor{white}{\textbf{Author (Partner)}} &
							\textcolor{white}{\textbf{Change description}}\\ \hline
1.0 & 26.02.2010 & Ismael Rivera (NUIG) & Final version ready for external review \\ \hline
\end{tabular}
\end{small}

@@

These back-end services cannot be directly used; hence they need to be encapsulated in what in the FAST platform is called Resource Adapters. The application of semantics to the back-end, through the corresponding resource adapters, and front-end building blocks assures a powerful instrument in the task of building new gadgets, improving the search, and enhancing the connection among the different building blocks which compose a gadget.

@@

\begin{tabular}
	{| >{\columncolor{fast@lightgrey}}p{3.25cm}|p{1.4cm}|p{3.28cm}|p{1.6cm}|p{3.29cm}|}
	\hline
	\textcolor{white}{\textbf{Delivery data}} & {\textbf{Due date}} & {28/02/2009} & {\textbf{Submitted}} & {27/02/2009}\\ \hline
	\textcolor{white}{\textbf{Status}} & \multicolumn{2}{l|}{} & \multicolumn{2}{l|}{final}\\ \hline
	\textcolor{white}{\textbf{Dissemination Level}} & \multicolumn{4}{l|}{Public $\boxtimes$ / Consortium $\square$}\\ \hline
	\textcolor{white}{\textbf{Short description of contents}} & \multicolumn{4}{p{10.85cm}|}{This deliverable presents techniques to exploit different kind of web services such as RESTful, SOAP-based and WSMO services. It exposes how these services are commonly defined, and what FAST need to allow these web services to be use within a gadget. In this deliverable is also presented a tool which apply some of these techniques allowing users to create the necessary wrappers for these services.}\\ \hline
	\textcolor{white}{\textbf{Authors}} & \multicolumn{4}{l|}{\authorOne}\\
	{} & \multicolumn{4}{l|}{\authorTwo}\\ 
%	{} & \multicolumn{4}{l|}{}\\ 
%	{} & \multicolumn{4}{l|}{}\\
	\hline
	\textcolor{white}{\textbf{Deliverable Owner}} & \multicolumn{2}{l|}{\authorOne} & \textbf{email} & {ismael.rivera@deri.org} \\ \cline{4-5}
	\textcolor{white}{\textbf{(Partner)}} & \multicolumn{2}{l|}{} & \textbf{phone} & {+353 91 495338} \\ \hline
	\textcolor{white}{\textbf{Keywords}} & \multicolumn{4}{p{10.85cm}|}{FAST, web service, WSDL, REST, SOAP, WSMO, Resource Adapter}\\ \hline
\end{tabular}
\end{small}
\newpage

@@

This deliverable is structured as follows. Section~\ref{sec:introduction} states the goal, scope and structure of the document. Section~\ref{sec:resource_adapters} describes what a resource adapter is, how the platform is able to interact with web services, how they these wrappers or adapters are build and then discovered to be reused in any gadget. Then, Section~\ref{sec:restful_web_services_wrapper_tool} defines the service wrapper tool built to facilitate FAST users to create resource adapters of RESTful web services. Finally, Section~\ref{sec:related_work}...\todo{Finish it!}

@@

\clearpage
\section{Resource Adapters} % (fold)
\label{sec:resource_adapters}

@@

The following sections explain what a resource adapter is, how a user, in this case a resource developer, can build these resource adapters, and how another user, a screen developer for instance, is able to discover and connect them while creating a screen. Hence, these three phases are called: building, discovery and connection.

\subsection{Definition} % (fold)
\label{sub:definition}

A complex gadget in terms of FAST is aimed to provide a functional access from a graphical user interface to a set of services (SOAP or REST-based Web services) and data sources (Atom/RSS feeds). These services and data sources need to be modelled and encapsulated within the platform in order to allow the discovery and use of them by other components (i.e. forms and operators). This is called Resource Adapter in the FAST platform. Figure~\ref{fig:complex_gadget_architecture} illustrates the role they play and which other building blocks they interact with within a gadget.

\begin{figure}[!htb]
  \begin{center}
    \includegraphics[width=\linewidth]{images/ComplexGadgetArchitecture.pdf}
    \caption{Complex Gadget Architecture}
    \label{fig:complex_gadget_architecture}
  \end{center}
\end{figure}

Basically, a resource adapter, from a design point of view, is composed of:

\begin{enumerate}
	\item \textbf{metadata} such as creator, label, description, version, and so on,
	\item a set of \textbf{actions} or operations,
	\item and a set of \textbf{postconditions} or outputs of the resource.
\end{enumerate}

Figure~\ref{fig:resource_adapter_overview} shows the information above in a graphical manner. An action (represented by a white square) is the minimum unit for the functionality a resource may have. Imagine a user clicking on a button to get a list of items for sale, or a checkout action performed in order to buy one of the items from the list. As an analogy, an action is what a operation is in a SOAP-based web service. Moreover, as these operations require some input, the actions may need a certain precondition (green circles) to be satisfied. And at last, the execution of these action may cause a postcondition (red circle) be satisfied, or following the same example, an output is returned as result of the execution of an operation.

\begin{figure}[!htb]
  \begin{center}
    \includegraphics[width=12cm]{images/resource_adapter.pdf}
    \caption{Resource Adapter Overview}
    \label{fig:resource_adapter_overview}
  \end{center}
\end{figure}

% subsection definition (end)

@@

This section describes different approaches for describing and implementing web services, and how a resource adapter could be generated for these services.

@@

REST \cite{Fielding2000} is a term to describe an architecture style, not a standard, of networked systems. The acronym REST stands for Representational State Transfer. REST-based or RESTful web services are created identifying all of the conceptual entities or resources which want to be exposed as services. Those entities or resources should be nouns, not verbs (orders, tickets, etc.). Then, the interaction with those resources is made by convention using HTTP verbs such as GET, POST, PUT or DELETE, in order to retrieve, create, modify or delete them.

RESTful web services are more lightweight than SOAP-based web services, avoiding the verbose XML-based messages, but unlike the former, which have a standard vocabulary, and commonly used, to describe the web service interface through WSDL, RESTful web services are currently not formally described most of the times. For a service consumer to understand the context and content of the data that must be sent to and received from the service, both the service consumer and service producer must have an out-of-band agreement. This takes the form of documentation, sample code, and an API that the service provider publishes for developers to use. For example, the many web-based services available from Google, Yahoo, Flickr, Amazon, and so on have accompanying artifacts describing how to consume the services. This style of documenting REST-based web services is fine for use by developers, but it averts tools from programmatically consuming such services and generating executable stubs for a given programming language, as a web service described using WSDL allows. Nevertheless, Web Application Description Language (WADL) attempts to resolve some of these issues by providing a means to describe services in terms of schemas, HTTP methods, and the request or response structures exchanged, but this language is not widely adopted yet by RESTful web service developers.

To permit a high number of REST-based web services to be integrated to the FAST platform, the approach taken is from a manual development perspective. There is no need of a formal document such as WADL defining the web service, for this reason, building service wrappers for these services involve the correct understanding of the service by a human being and a tool to facilitate this task. This tool is explained in detail in Section~\ref{sec:restful_web_services_wrapper_tool}.

@@

SOAP-based web Services or "Heavyweight Web Services" use Extensible Markup Language (XML) \cite{XML1.1} messages that follow the Simple Object Access Protocol (SOAP) standard \cite{SOAP1.2} and have been popular within traditional enterprise, usually relying on HTTP for message negotiation and transmission. In such services, there is often a machine-readable description of the operations offered by the service written in a Web Services Description Language (WSDL) document. The advantage of specifying the service in a using WSDL is that it can be programmatically processed.

@@

From a specific WSDL definition, a set of methods or operations can be easily extracted which encloses a set of inputs and outputs. These operations would be transformed into \emph{actions}, and the inputs and outputs would be used to define resource adapter \emph{pre/postconditions}. But there is a critical problem following this approach. The definition of the inputs, outputs, operations and their types are done in a XML Schema document. An XML schema describes, in a syntactic manner, the structure of an XML document, but there is no straight-forward relation with the concepts used to describe pre/postconditions in the building blocks. In order to illustrate the concepts of WSDL, and the gap between them and semantically-defined building blocks, a shorten version of a real Amazon E-commerce web service will be presented. Concretely, the choosen web service performs an item search through the Amazon catalog. It expects as input an ItemSearch element, which may contain an actor, an artist or an author, and the item availability. The output of the service would be a set of items. The WSDL including for this service is given below.

\singlespacing
\begin{small}
\verbatiminput{examples/amazon.wsdl}
\end{small}
\doublespacing

As stated before, one of the advantages of describing web services in a formal document is that they can be programmatically consumed. In fact, there are many frameworks out there to facilitate such a task, parsing WSDL files, creating stubs to consume the web service and so on. However, in order to create a resource adapter from the WSDL defined above, the types defined in the XML Schema need to be translated into concepts from an ontology.

This problem could be addressed in two different ways. First one, as a manual process mapping these types and their attributes to concepts and properties, as it was proposed for RESTful services. Second approach is taking advantage of SAWSDL, which keeps the process in the programmatically direction.

SAWSDL \cite{sawsdl} stands for Semantic Annotations for WSDL and XML Schema, and it defines how to add semantic annotations to various parts of a WSDL document such as input and output message structures, interfaces and operations. These semantic annotations enable the possibility to develop resource adapters from WSDL documents. They reference a concept in an ontology or a mapping document, and are either independent of the ontology expression languages and mapping languages.

Basically, there are three main concepts about SAWSDL: \emph{modelReference} to specify the association between a WSDL or XML Schema component and a concept from the ontology, and \emph{liftingSchemaMapping} and \emph{loweringSchemaMapping} to define the mapping mechanisms between semantic data and XML. These lowering and lifting schema mappings are not explained in this deliverable, however in Appendix A of the SAWSDL specification \cite{sawsdl} they are described in detail, and illustrated with examples.

To illustrate the usage of these attributes, the following code shows a fragment of the previous example using SAWSDL to semantically annotate the service with concepts from an imaginary ontology defining the Amazon E-commerce web service.

\singlespacing
\begin{small}
\verbatiminput{examples/amazon_sawsdl.wsdl}
\end{small}
\doublespacing

Hence, SAWSDL enriches the web services descriptions made in WSDL with references to concepts from a semantic model and enables the possibility to create resource adapters from these definitions without a strong user participation in the process. If the semantic metadata is missing, a manual annotation process could be considered, where the user would need to relate concepts from an ontology with the WSDL or XML Schema elements, and create the mapping between them manually as well.

@@

A web service in WSMO defines one and only one capability. The capability of a web service defines its functionality in terms of pre/postconditions, assumptions and effects. Apart from that, a web service capability is defined by specifying the following elements: non-functional properties, imported ontologies, used mediators, and shared variables. The creation of the resource adapters for these web services will just take into consideration the set of pre/postconditions.

In the rest of the section, the Amazon E-commerce web service will be used to build a WSMO web service with the same functionality \cite{wsmo_amazon}. In fact, just the small subset of the service, used in previous sections, will be tranformed.

First of all, an ontology needs to be extracted from the XML Schema types of the WSDL definition. For the shake of simplificity, this example does a one-to-one mapping between the Schema types and the concepts, though a real sceneario may consider an different approach to optimise the modelling.

The concept describing an item search request for the Amazon web service is:

\singlespacing
\begin{small}
\begin{verbatim}
concept itemSearchRequest
    nonFunctionalProperties
      dc#description hasValue "A request to search for an item"
    endNonFunctionalProperties
    author ofType _string
    title ofType (1) _string
    rating ofType ratings#audienceRating
    minPrice ofType (1) _decimal
    maxPrice ofType (1) _decimal
\end{verbatim}
\end{small}
\doublespacing

The concept \emph{itemContainer} describes a list of items returned in the item search response, and its code is presented below:

\singlespacing
\begin{small}
\begin{verbatim}
concept itemContainer
    nonFunctionalProperties
      dc#description hasValue "Contains a list of items"
    endNonFunctionalProperties
    items ofType item
\end{verbatim}
\end{small}
\doublespacing

Next step would be to describe the namespace and non-functional properties declarations for the WSMO Amazon web service, but as it was said, this information is not relevant for the example. Hence, the simplified version of the Amazon web service preconditions would look like:

\singlespacing
\begin{small}
\begin{verbatim}
precondition
  nonFunctionalProperties
     dc#description hasValue "The Amazon service handles requests 
     for searching for items by keywords"
  endNonFunctionalProperties
  definedBy
    // A search by author, title, minPrice or maxPrice
    ?request memberOf am#itemSearchRequest
\end{verbatim}
\end{small}
\doublespacing

It is a fairly simple precondition, which the only condition required is the existence of a \emph{itemSearchRequest} object.

However, postconditions are a bit more difficult to construct. They take the form of implication rules, so given a particular precondition, one particular postcondition will take effect. To follow the same example, these would be the postconditions for the Amazon search items functionality, including just the criterion author, title, minimum and maximum price as defined in the ontology.

\singlespacing
\begin{small}
\begin{verbatim}
 postcondition
  nonFunctionalProperties
    dc#description hasValue "The service returns a list of searched
      items"
  endNonFunctionalProperties
  // The result for a search request by author is a container of products
  // such that all items in the container have the requested author
  (?request[
      author hasValue ?author
    ] memberOf am#searchItems implies 
    exists ?container 
      (?container memberOf am#itemContainer and 
        forall ?item 
          (?container[
              items hasValue ?item
            ] implies 
            ?item[
              author hasValue ?author
            ]
          )
      )
  ) and

  // The result for a search request by title is a container of products
  // such that all items in the container have the requested title
  (?request[
      title hasValue ?title
    ] memberOf am#searchItems implies 
    exists ?container 
      (?container memberOf am#itemContainer and 
        forall ?item 
          (?container[
              items hasValue ?item
            ] implies 
            ?item[
              title hasValue ?title
            ]
          )
      )
  ) and

  // The result for a search request by minPrice is a container of products
  // such that all items in the container have a price >= requested minPrice
  (?request[
      minPrice hasValue ?minPrice
    ] memberOf am#searchItems implies 
    exists ?container 
      (?container memberOf am#itemContainer and 
      forall{?item,?price} 
        (?container[
            items hasValue ?item
          ] and 
          ?item[
            price hasValue ?price
          ] implies 
          ?price >= ?minPrice
        )
      )
  ) and

  // The result for a search request by maxPrice is a container of products
  // such that all items in the container have a price <= requested maxPrice
  (?request[
      maxPrice hasValue ?maxPrice
    ] memberOf am#searchItems implies 
    exists ?container 
      (?container memberOf am#itemContainer and 
      forall {?item,?price} 
        (?container[
            items hasValue ?item
          ] and 
          ?item[
            price hasValue ?price
          ] implies 
          ?price =< ?maxPrice
        )
      )
  ).
\end{verbatim}
\end{small}
\doublespacing

The manner pre/postconditions are defined in WSMO web services are similar to the way they are defined in the Resource Adapters. Section 4.4 in \cite{moller2010fast_ontology} gives a detailed explanation on this topic. Basically, and using previous examples, the preconditions would be transformed as follows:

\begin{verbatim}
?request memberOf am#itemSearchRequest => ?request rdf:type am:itemSearchRequest
\end{verbatim}

And a simple postcondition would be transformed as follows:

\begin{verbatim}
exists ?container 
      (?container memberOf am#itemContainer) => ?container rdf:type am:itemContainer
\end{verbatim}

@@

\subsection{Discovery} % (fold)
\label{sub:discovery}

In the building phase, every resource adapter is created sharing a common structure for screen components, such as forms, operators and resources. Hence, it will have a set of actions which will contain a set of preconditions to satisfy in order to be executed, and after its execution, it may produce any of the conditions of their postconditions.

@@

On these basis, the mechanism is relatively similar to the screen recommender, used to find the best screens to create a screen flow. Once more, it is based on the pre/postconditions of a set of building blocks.

@@

\subsection{Connection} % (fold)
\label{sub:connection}

A resource adapter is aimed to perform some action on a web service. In order to perform a given action, a precondition may need to be fulfilled, and as a result of the execution a postcondition may be generated. These pre/postconditions are intended to be used by pre/postconditions of other building blocks within a gadget. This connection must be explicitly defined, because it will indicate the real data-flow between different building blocks. In FAST this is commonly called Piping, since a pipe in computer science is just that, a one-way communication channel for interprocess communication; therefore a pipe is a logical connection made between a precondition and a postcondition of different building blocks. Figure~\ref{fig:Piping} shows an example of a search form for the Amazon service. This is a screen, composed by a form and a resource adapter. The form allows the user to introduce a term for the search, producing a postcondition which will be redirect to a resource adapter precondition. The execution of the web services produces a list of results, which will be forward to the postcondition of the screen.

\begin{figure}[!htb]
  \begin{center}
    \includegraphics[width=12cm]{images/piping.pdf}
    \caption{Piping}
    \label{fig:Piping}
  \end{center}
\end{figure}

% subsection connection (end)

@@

In FAST, the wrapper tool will be in charge of the resource adapters' building phase. At the moment, it just supports RESTful web services. The building phase for this type of web services is done in two steps: a first step will be in charge of the construction of a service request and a second step will analyse the response got from the execution of the service, allowing the extraction of facts contained in that response and mapping them to domain-specific concepts from the ontologies used within FAST by any building block.

@@

To achieve this, service wrappers will be handled as any buis with preconditions and postconditions (inputs and outputs). Therefore, these precondition ports might be used to determine values to parameters like query keywords coming from a building block inside the screen the service is placed or even an external screen as a screen precondition. The service wrapper tool composes the request using a template string which will contain placeholders for precondition values. Before sending the request to the service, the placeholders are replaced with their corresponding values and the resulting URL is then ready to be sent the service request.

To achieve this, service wrappers will be handled as any other building block with preconditions and postconditions (inputs and outputs). Therefore, these precondition ports might be used to determine values to parameters like query keywords coming from a building block inside the screen the service is placed or even an external screen as a screen precondition.

To define the preconditions and postconditions of a new service wrapper, the FAST service wrapper tool provides a form allowing the edition of these entries, cf. Figure~\ref{fig:construct_pre_post_conditions}\footnote{Note: These example values may be used to test the service wrapper, but they can be freely edited.}

\begin{figure}[!htb]
  \begin{center}
    \includegraphics[width=\linewidth]{images/ServiceWrapperToolGVSWithRequestExample.jpg}
    \caption{Configuring pre/postconditions of a service wrapper}
    \label{fig:construct_pre_post_conditions}
  \end{center}
\end{figure}

@@

Figure~\ref{fig:construct_service_request} shows the screen to construct the service requests. In the top input field of Figure~\ref{fig:construct_service_request}, the user may drop an example HTTP request taken e.g. form the service documentation (e.g. \cite{eBayShoppingAPIs} for this example). The tool analyses the example request and, in the middle of the screen, a form for editing the request parameters is provided. This example shows how a user has connected the \textit{QueryKeywords} parameter with the precondition \textit{search\_key} by adding a corresponding reference to the value field of that parameter. In addition, an access key has been retrieved and has been entered as value for the \textit{appid} parameter.

\begin{figure}[!htb]
  \begin{center}
    \includegraphics[width=\linewidth]{images/ServiceWrapperToolGVSWithRequestExample.jpg}
    \caption{Constructing the service request URL and its parameters}
    \label{fig:construct_service_request}
  \end{center}
\end{figure}

Below the request paramter editing form of Figure~\ref{fig:construct_service_request}, a \textit{Send Request} button allows to validate the constructed service by sending it to the specified service address (via a server relay). Then the placeholders for preconditions are replaced with example values and the resulting HTTP request is shown below the parameter form. In addition, the request is sent and the response is shown on the bottom of that page, which helps the users to figure out quickly whether the constructed request works as desired.

@@

Once the service request is constructed and sent to the service provider, it will send back a response. This response message could be serialize in any format, though the most common formats used nowadays are XML or JSON among others. To continue the example it was started in the previous section, the response of the Ebay Shopping service will be in XML as specified in the request, which is the format supported by the wrapper tool.

Figure~\ref{fig:response_service_execution} shows the transformation data tab of the wrapper tool.

\begin{figure}
  \begin{center}
    \includegraphics[angle=90,width=0.8\linewidth]{images/ServiceWrapperToolGVSWithTransformationRules.jpg}
    \caption{Interactive rule based transformation of an XML response to FAST facts}
    \label{fig:response_service_execution}
  \end{center}
\end{figure}
